Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:  

- diagonal line == accuracy line
- people tend to guess 'round' numbers: 50, 100, 200
- people underestimate their audience

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
library(ggthemes)
theme_set(theme_minimal())
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point()

```

***

#### What are some things that you notice right away?
Response:

- younger people seems to have more friends (and definitely more high-friend outliers)
- age 70 (69?) looks strongly overrepresented
- ages > 100 looks strongly overrepresented

***

### ggplot Syntax
Notes:

- ggplot2 tutorial: http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf
- list of geoms: http://docs.ggplot2.org/current/
- Grammar of Graphics (book behind ggplot2 grammar): http://www.amazon.com/The-Grammar-Graphics-Statistics-Computing/dp/0387245448

```{r ggplot Syntax}
```

***

### Overplotting
Notes:

```{r Overplotting}
# to find out min and max of pf$age
# summary(pf$age)
# 13.00   20.00   28.00   37.28   50.00  113.00 

# also, setting alpha = 1/20 - it takes 20 points to build to 'full' colour point.
# geom_jitter (instead of geom_point) - at random noise to disperse the vertical columns 
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 1/20) + 
  #scale_y_log10() + 
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:

- younger users friedn count isn't much higher than older ones - it's just many high value outliers which creates such impression
- younger users have more friends than olders ones
- younger users have more high value outliers
- there is more younger users (?)
- distribution is looking a bit bimodal with values decreasing ~40 and going up ~50
***

### Coord_trans()
Notes:
Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r Coord_trans()}
# Dealing with jitter values < 0
#ggplot(aes(x = age, y = friend_count), data = pf[pf$friend_count>0,]) +
#  geom_jitter(alpha = 1/20) + 
#  coord_trans(ytrans = "sqrt") +
#  xlim(13, 90)

ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 1/20, position = position_jitter(h = 0)) + 
  coord_trans(ytrans = "sqrt") +
  xlim(13, 90)
```


#### What do you notice?
There is more younger users of facebook (density of points for age < 30 is higher than for other ages)

***

### Alpha and Jitter
Examine the relationship between friendships_initiated (y) and age (x) using the ggplot syntax.

We recommend creating a basic scatter plot first to see what the distribution looks like and then adjusting it by adding one layer at a time.

Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_jitter(alpha = 1/20, position = position_jitter(h = 0)) +
  scale_y_sqrt() +
  xlim(13, 90)
```

***

### Overplotting and Domain Knowledge
Notes:

- choose proper (sometimes transformed) values for x and y.
  e.g. Moira changed the x and y scales from 'how many friends saw your post' to '% of friends saw your post'.
  This scales nicely with # of friends people have.

***

### Conditional Means
More dplyr resources:

- Introduction to dplyr: http://rstudio-pubs-static.s3.amazonaws.com/11068_8bc42d6df61341b2bed45e9a9a3bf9f4.html
- Hadley Wickham "dplyr" talk:
    - http://www.r-bloggers.com/hadley-wickham-presents-dplyr-at-user-2014/
    - http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/
    - http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-2/
- lapply blog post: https://rollingyours.wordpress.com/2014/10/20/the-lapply-command-101/

```{r Conditional Means}
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_by_age = summarise(age_groups, 
                         friend_count_mean = mean(friend_count),
                         friend_count_median = median(friend_count),
                         n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
```

```{r Conditional Means alternate code}
pf %>% 
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)
```

Create your plot!

```{r Conditional Means Plot}
head(pf.fc_by_age, 20)
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  geom_bar(stat = 'identity', color='black', fill='#02a0e2')

ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  #geom_point() +
  geom_line(stat = 'identity', color='#02a0e2')
```

***

### Overlaying Summaries with Raw Data
We take the jittered friendship graph as the base and add 4 summary overlays on top of it:

- mean (black line)
- median (blue line)
- 10th and 90th percentile (blue dashed lines)

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  #xlim(13, 90) +
  #coord_trans(ytrans = "sqrt") +
  coord_cartesian(xlim = c(13, 90), ylim=c(0, 1000)) +
  geom_point(alpha = 0.05, 
             position = position_jitter(h = 0),
             color = 'orange') + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = median, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1, linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9, linetype = 2, color = 'blue')

```

#### What are some of your observations of the plot?
Response:

- there is a high peak of friends in young users group (<25) which then flats ~100 for older users
- having > 1000 friends is rare
- most (90%) users with age > 30 have less than 250 friends
- age = 69 and age > 80 seems like fake age data - they have large friends count, similar to values of youngest users
- median and percentile lines follow the same trend as mean - there are no anomalies in 10% of lowes or highest friend counts
***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

- histograms are used to show quntitative data about how user perceive their audience

***

### Correlation
Notes:

- visual guide to correlation: https://s3.amazonaws.com/udacity-hosted-downloads/ud651/correlation_images.jpeg
- correlation coefficient: http://www.r-tutor.com/elementary-statistics/numerical-measures/correlation-coefficient
- udacity correlation: https://www.udacity.com/course/viewer#!/c-ud201/l-1345848540/m-171582737
- different methods of correlation: http://www.statisticssolutions.com/correlation-pearson-kendall-spearman/

```{r Correlation}
cor.test(pf$age, pf$friend_count, method="pearson")
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:
cor = -0.027

Results:
-0.3 < cor < 0.3 - no correlation  
0.3 < cor < 0.5 - small correlation  
0.5 < cor < 0.7 - moderate correlation  
0.7 < cor - large correlation  

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(pf[pf$age<=70,] , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes: correlation numbers are useful but do not give as much insight as plotting the data.

***

## Create Scatterplots
Notes:

```{r}
# There are few significant outliers that extend the axis. Let's remove them as 1-2% top outliers. Where to cut?
# quantile(pf$likes_received, probs = c(.90, .95, .98))

ggplot(aes(x = www_likes_received, y = likes_received), data=pf) +
  geom_point(alpha = 0.05) +
  coord_cartesian(xlim = c(1, 600), ylim = c(1, 600)) +
  scale_x_sqrt() +
  scale_y_sqrt() + 
  # add correlation line
  geom_smooth(method = 'lm', color = 'red')
```

***

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
cor.test(pf$www_likes_received, pf$likes_received, method="pearson")
```

Response: 0.948

***

### Moira on Correlation
Notes:

- be aware of the correlation of your variables (e.g. # of status updates and # of photos uploaded).
  Such correlations might describe same behavior/situation (here: how user is engaged). Later on, as you try to 
  build models, you should choose if you want to keep both strongly correlated vars as model input.

***

### More Caution with Correlation
Notes:

- dataset of soil temp per month (one observation per month). Month values goes from 1-203
- examples about tricky correlation; book Applied Linear Regression: http://users.stat.umn.edu/~sandy/alr3ed/website/

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
data(Mitchell)

ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

0.05

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
cor.test(Mitchell$Month, Mitchell$Temp, method="pearson")
```

0.057

***

### Making Sense of Data
Break up the x-axis in 12 months increments 

```{r Making Sense of Data}
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point() + 
  scale_x_discrete(breaks = seq(0, 203, 12))
```

***

### A New Perspective
Zoom in into the plot, stretch y-axis and extend the x-axis. 

What do you notice?  
Response:  
Every year, temperature follows the same pattern: starts with low values on January, goes up
and peaks around June, July. Then, the temperature symmetrically goes down to lowest values in December.
Function looks like first half of sinusoidal function.

Watch the solution video and check out the Instructor Notes!
Notes:

- make sure you use whole context about the data you have. Here: months are cyclic
- use proper scales for the plot: here, with x extended and y stretched, the cyclicity of data could be seen
- William Playfair and the Psychology of Graphs: http://www.psych.utoronto.ca/users/spence/Spence%20(2006).pdf
- there are other tests that can find the cyclic correlation in data (e.g. library(energey); dcor.ttest())

```{r}
ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point() 

```

***

### Understanding Noise: Age to Age Months
Notes: concrete values of means are just sample of the data and still contains some noise inside them.
The effect is event more visible if we count mean_friend_count / age_by_month.

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  geom_line()
head(pf.fc_by_age, 10)
pf.fc_by_age[17:19,]
```

***

### Age with Months Means

```{r Age with Months Means}
# Use December 31 as the reference date.
pf$age_with_months = pf$age + (12 - pf$dob_month)/12
```

Programming Assignment
```{r Programming Assignment}
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
             n = n()) %>%
  arrange(age_with_months)
head(pf.fc_by_age_months)

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
ggplot(aes(x=age_with_months), data=pf.fc_by_age_months[pf.fc_by_age_months$age_with_months<71,]) +
  geom_line(aes(y = friend_count_mean))
#  geom_line(aes(y = friend_count_mean), color='#109618') +
#  geom_line(aes(y = friend_count_median), color='#cc3637')

```

***

### Smoothing Conditional Means
Notes: 

- by decreasing the bucket size (age -> age_by_month) we increase the noise of the plot.
- plots below represent variance/bias tradeoff (e.g. miss features (age 69))

Local Regression:

- explained visually: http://simplystatistics.org/2014/02/13/loess-explained-in-a-gif/
- Simply Statistics blog: http://simplystatistics.org/
- local regression wikipedia: https://en.wikipedia.org/wiki/Local_regression

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x = age, y = friend_count_mean), data = subset(pf.fc_by_age, age < 71)) +
  geom_line() +
  geom_smooth()
  
p2 <- ggplot(aes(x = age_with_months, y = friend_count_mean), data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()
# smoothened mean
p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count), data = subset(pf, age<71)) + geom_line(stat = 'summary', fun.y = mean)

library(gridExtra)
grid.arrange(p2, p1, p3, ncol = 1)

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

- Lesson covered:
    - scatterplots
    - conditional means
    - correlation
    - noise
- scatterplots:
    - main tool for analysis of two variables
    - great tool to reveal outliers and anomalies in data (age 69, different variance for young and old users)
    - with proper layers (jitter, mean, median, qunatiles), one scatterplot presents clearly lots of information
- conditional means:
    - augments scatterplots
    - perfect for showing trends
    - makes me thing about groups of data (here: age) and how to split it
- correlation
    - quick test of hypothesis (is #of friends going down with age?)
    - you need to understand the data before (correlation of soil temp vs month)
- noise
    - varianve vs bias
    - use smoothening to remove the noise and find  trends

Extra notes:

- A deep dive into Bivariate Data Analysis: http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat401/Notes/401-bivariate-slides.pdf
    - bivariate = two parameters for each simple random sample
    - marginal property = property that describes X values only or Y values only
    - joint property = property that describex X and Y at the same time
    - joint distribution = complete summary of stat props of (X, Y)
          - joint distribution is not described completelty by marginal distribution of both variables
    - scatterplot = most important graphical summary of data
    - correlation
          - Pearson
                - r == coefficient (Pearson correlation coefficient)
                - covariance (numerator of r)
                - page 21 and 22 shows chart how values contribute to coefficient value
    - ... more

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

