Problems set 4
============================

Price vs. x
-------------

```{r}
library(ggplot2)
library(ggthemes)
theme_set(theme_minimal())
library(dplyr)
data(diamonds)

summary(diamonds$x)
ggplot(aes(x=x, y=price), data=diamonds) +
  geom_point(alpha = 0.05) + 
  scale_x_continuous(limits=c(3,10), breaks=seq(3,10,0.5))

```

Observations:

- price seems to grow exponentially wit the lenght (x) value
- there seems to exist 'size clusters': e.g. for value of ~5.25 there is significantly more observations than for 5.0 and 5.5;
  same for larges sizes (e.g. 5.75)
- there is less large diamonds (size > 7) than small and medium ones
- one can't buy small diamonds for more than 5000$
- buying large diamond for < 5000$ is almost impossible

```{r}
# correlation between price and x (0.88)
cor.test(diamonds$price, diamonds$x)

# correlation between price and y (0.87)
cor.test(diamonds$price, diamonds$y)

# correlation between price and z (0.86)
cor.test(diamonds$price, diamonds$z)
```

Price vs. depth
---------------
```{r}
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point(alpha = 0.1) +
  scale_x_continuous(limits=c(50, 70))

# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units.
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point(alpha = 0.01) +
  scale_x_continuous(limits=c(50, 70), breaks=seq(50, 70, 2))
```

Typical depth range: 58 to 64.

```{r}
# What's the correlation of depth vs price? (-0.01)
cor.test(diamonds$depth, diamonds$price)
```

Would you use depth to predict price of a diamond: No
Why: low absolute value of correlation coefficient (next to 0) means there is no correlation between depth and price.
Diamonds of any depth can achieve both very low and very high price.

Price vs. carat
-----------------
```{r}
diamonds.top1p_excluded = subset(diamonds, 
                                 diamonds$carat < quantile(diamonds$carat, .99) & diamonds$price < quantile(diamonds$price, .99))
ggplot(aes(x = carat, y = price), data = diamonds.top1p_excluded) +
  geom_point()

```

Price vs. volume
----------------
```{r}
diamonds$volume = diamonds$x * diamonds$y * diamonds$x
ggplot(aes(x = volume, y = price), data = diamonds) +
  geom_point(alpha = 0.05) +
  scale_x_log10(limits=c(50, 1000), breaks=c(50, 100, 200, 300, 400, 500, 700, 1000))
```

Observations:

- price vs volume follows similar trend as price vs x (exponential growth of price with x)
- there are volume clusters (e.g. 150, 200, 400, 500)
- diamonds with volume < 350 are most popular
- log scale is not required here - continuous scale works better (easier to interpret)

Did not notice:

- outliers:
    - expensive diamond near 4000 and a cheaper diamond with a volume near 900
      (I removed the outliers to make the chart x-axis limits better)
- diamonds with volume == 0 (8!)

```{r}
# What's the correlation of price and volume? (0.92)
# NOTE: instructor notes suggest removing outliers (vol > 800 && vol = 0)
cor.test(diamonds$price, diamonds$volume)
```

### Adjustments - price vs. volume
```{r}
# Subset the data to exclude diamonds with a volume
# greater than or equal to 800. Also, exclude diamonds
# with a volume of 0. Adjust the transparency of the
# points and add a linear model to the plot. (See the
# Instructor Notes or look up the documentation of
# geom_smooth() for more details about smoothers.)
require(mgcv)
diamonds.vol_nooutliers = diamonds[diamonds$volume < 800 & diamonds$volume != 0, ]
ggplot(aes(x = volume, y = price), data = diamonds.vol_nooutliers) +
  geom_point(alpha = 0.05) + 
  stat_smooth(method = 'lm', formula = y ~ x, color = 'brown', size = 1) +
  stat_smooth(method = 'lm', formula = y ~ poly(x, 2), color = 'red', size = 1) +
  stat_smooth(method = 'lm', formula = y ~ poly(x, 3), color = 'blue') +
  stat_smooth(method = 'gam', formula = y ~ s(x), color = 'green')
```

Observations:

- linear and square models are worse than cubic and gam models - they overprice smaller diamonds
- estimating value of diamonds with volume > 300 is not doable with any models - variance is too high and model
  cannot determine the price based on volume only

Mean price by clarity
---------------------

Use the function dplyr package to create a new data frame containing info on diamonds by clarity.

Name the data frame diamondsByClarity

The data frame should contain the following variables in this order.

    1. mean_price
    2. median_price
    3. min_price
    4. max_price
    5. n
    
```{r}
diamondsByClarity = group_by(diamonds, clarity)
diamondsByClarity = summarise(diamondsByClarity,
                              mean_price = mean(price),
                              median_price = median(price),
                              min_price = min(price),
                              max_price = max(price),
                              n = n())
head(diamondsByClarity)
```


Bar Charts of Mean Price
------------------------
We’ve created summary data frames with the mean price
by clarity and color. You can run the code in R to
verify what data is in the variables diamonds_mp_by_clarity
and diamonds_mp_by_color.

Your task is to write additional code to create two bar plots
on one output image using the grid.arrange() function from the package
gridExtra.

flowigdata.com article on Bar Charts vs Histograms: http://flowingdata.com/2014/02/27/how-to-read-histograms-and-use-them-in-r/

```{r}
data(diamonds)
library(dplyr)
library(gridExtra)

diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

#diamonds_by_cut <- group_by(diamonds, cut)
#diamonds_mp_by_cut <- summarise(diamonds_by_cut, mean_price = mean(price))

p1 <- ggplot(aes(x = clarity, y = mean_price), data = diamonds_mp_by_clarity) + geom_bar(stat='identity')
p2 <- ggplot(aes(x = color, y = mean_price), data = diamonds_mp_by_color) + geom_bar(stat='identity')
#p3 <- ggplot(aes(x = cut, y = mean_price), data = diamonds_mp_by_cut) + geom_bar(stat='identity')
grid.arrange(p1, p2, ncol = 1)
```

Observations:

- color seems to define a scale with 'best' and 'worst' colors (price for J > ... > price for D)
- clarity also seems to define some type of order with spike on SI2 and lowest price values -> IF
- (instructor notes): mean price tends to decrease as clarity improve; the same can be said for color.  
  Look at mean price across cut.
  
Gapminder Revisited
-------------------
```{r}
library(openxlsx)
library(tidyr)
fertility <- read.xlsx('indicator undata total_fertility.xlsx', rowNames = TRUE, rows=1:259)
fertility <- mutate(fertility, country=rownames(fertility))
fertility <- gather(fertility, year, "fertility", 1:(length(fertility)-1), convert=TRUE)
fertility <- arrange(fertility, country, year)

income <- read.xlsx('indicator gapminder gdp_per_capita_ppp.xlsx', rowNames = TRUE, rows=1:260)
income <- mutate(income, country=rownames(income))
income <- gather(income, year, "fertility", 1:(length(income)-1), convert = TRUE)
income <- arrange(income, country, year)

fer_inc <- inner_join(income, fertility, by = c("country", "year"))
names(fer_inc) <- c("country", "year", "income", "fertility")
head(fer_inc)
```

### How did fertility income change in time?
```{r}
summary(fer_inc$income)
ggplot(aes(x = year, y = income), data = fer_inc) +
  geom_point(alpha = 0.2, color = 'orange') +
  coord_cartesian(ylim=c(300, 190000)) +
  scale_y_log10() +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = median, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1, linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9, linetype = 2, color = 'blue')

fer_inc[!is.na(fer_inc$income) & fer_inc$income < 300,]
```

Observations:

- there are some values of income < 300 (Liberia in 1992-1997) which were removed from plot for better axis adjustment.
  What happened in Liberia in 1990s? [civil war]
- there are countries with income > 100000 (Brunei, Kuwait, Macao, Qatar and United Arab Emirates). It's good to have lots
  of oil, right?:)  
- general trend is that income is growing (significantly!); the growth started in 1900 and continues until now
- median and mean do not go far from each other.  
  Probably the income distribution is ~ normal distribution

### How did income and fertility changed in time?
```{r}
fer_inc.means <- fer_inc %>%
  group_by(year) %>%
  summarise(mean_income = mean(income, na.rm = TRUE),
            mean_fertility = mean(fertility, na.rm = TRUE),
            n = n()) %>%
  arrange(year)
head(fer_inc.means)
p1 <- ggplot(aes(x = year), data = fer_inc.means) +
  geom_point(aes(y = mean_income), color = 'red')
p2 <- ggplot(aes(x = year), data = fer_inc.means) +  
  geom_point(aes(y = mean_fertility), color = 'blue')
library(gridExtra)
grid.arrange(p1, p2, ncol=1)

cor.test(fer_inc$income, fer_inc$fertility)
```

Observations:

- as the mean income is going up, so does the mean fertility rates are dropping. 
  It seems like both value are correlated (Pearson: -0.48), so increase in income causes decrease in fertility.
  