Lesson 3
========================================================


### What to Do First?
Notes:
Read data and start with some basic look at the data: what is the format (head), what are the columns (head), what data do the columns have (str, summary).
***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
getwd()
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
head(pf)
summary(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
# install.packages('ggplot2')
library(ggplot2)
library(ggthemes)
theme_set(theme_minimal())
names(pf)
# qplot(x = dob_day, data = pf)
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram() + 
  scale_x_discrete(breaks = 1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
- 1st has noticeably more counts that any other data. This is most likely due to fact, that some people don't want to
  specify their birth date. To put sth in, they choose 1st
- similar (but smaller) effect can be seen for 10th and 20th - they both have small spikes.
- interesting, if the same situation happens for months (more January than any other) or years (any special year?)
***

### Moira's Investigation
Notes: people significantly underestimate their audience size (estimate is usually 25% of real value).

***

### Estimating Your Audience Size
Notes: --

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response: --

#### How many of your friends do you think saw that post?
Response: --

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response: --

***

### Perceived Audience Size
Notes: --

***
### Faceting
Notes:

```{r Faceting}
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram() +
  scale_x_discrete(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
# facet_wrap(formula)
# facet_wrap(~variable)
# use . to skip variable (http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/)
# facet_grid(vertical ~ horizontal) 
```

#### Let’s take another look at our plot. What stands out to you here?
Response: Jan 1st is significantly overrepresented. Other '1st' days of month do not present such a big anomaly anymore.

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes: histogram of estimated views had one outlier with estimated 10M views.
This has clealry skewed the histogram by extending the X-axis.


#### Which case do you think applies to Moira’s outlier?
Response: bad data about extreme case.

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram()
```

#### How is this plot similar to Moira's first plot?
Response: there are some huge outliers (friend_count ~5000) that extend the X-axis making it hard to
see what is happening with the bulk of data. 
This kind of distribution is called 'Long-tail' data.

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```

### Exploring with Bin Width
Notes: people tend to guess round values (10, 20, 30, 50, 100). Setting the bin_width = 1 helps to reveal this pattern.

***

### Adjusting the Bin Width
Notes: many people have < 25 friends (or no friends at all, after setting bin_width = 1)

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_grid(gender ~ .)
```

***

### Omitting NA Values
Notes: histogram do not answer the question clearly ('Who has more firends: men or women?').

```{r Omitting NA Values}
ggplot(aes(x = friend_count), data = subset(pf, !is.na(gender))) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)

# stats: who on average has more?
#y$female[4] - y$male[4] 
# stats: median difference
#y$female[3] - y$male[3] 
```

#### Who on average has more friends: men or women?
Response: women

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
ggplot(aes(x = tenure), data = pf) +
  geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
ggplot(aes(x = tenure/365), data = pf) +
  geom_histogram(binwidth = 0.25, color = 'black', fill = 'orange') +
  scale_x_continuous(limits = c(0, 8), breaks = seq(0, 8,1))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
ggplot(aes(x = tenure/365), data = pf) +
  geom_histogram(binwidth = 0.25, color = 'black', fill = 'orange') +
  scale_x_continuous(limits = c(0, 8), breaks = seq(0, 8,1)) +
  xlab('Number of years using Facebook') +
  ylab('Number of users in sample')
```

***

### User Ages
Notes:

```{r User Ages}
ggplot(aes(x = age), data = pf) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#5760AB') +
  #scale_x_continuous(limits = c(10, 120), breaks = seq(10, 113, 5)) +
  scale_x_discrete(breaks = seq(10, 113, 5)) +
  xlab('Age of Facebook user') +
  ylab('Number of users in sample')  
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes: Lara's site and Memes papers:
http://www.ladamic.com/
http://www-personal.umich.edu/~ladamic/papers/memes/howmet.pdf

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes: create a plot with 3 histograms: untransformed friend_count, log10 and sqrt

```{r Add a Scaling Layer}
summary(pf$friend_count)
summary(log(pf$friend_count + 1))
summary(sqrt(pf$friend_count + 1))

# create a plot with 3 histograms: untransformed friend_count, log10 and sqrt
fc_reg <- ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 50, color = 'black', fill = '#5760AB')
#fc_reg

fc_log <- ggplot(aes(x = log10(friend_count+1)), data = pf) +
  geom_histogram(binwidth = 0.1, color = 'black', fill = '#5760AB')
#fc_log

# alternatively, we can use scale_x_log()
#fc_log <- ggplot(aes(x = friend_count), data = pf) +
#  geom_histogram(binwidth = 0.1, color = 'black', fill = '#5760AB') +
#  scale_x_log10()
#fc_log

fc_sqrt <- ggplot(aes(x = sqrt(friend_count)), data = pf) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#5760AB')
#fc_sqrt

library(gridExtra)
grid.arrange(fc_reg, fc_log, fc_sqrt, ncol=2)
```

***


### Frequency Polygons

```{r Frequency Polygons}
ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('friend count') + 
  ylab('proportion of users with that friend count')
  
# you can try y = ..density..
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
ggplot(aes(x = www_likes, y = ..count..), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) +
  #geom_freqpoly(aes(color = gender), binwidth=0.5) +
  scale_x_continuous() +
  scale_x_log10() +
  xlab('likes count') + 
  ylab('proportion of users with that likes count')

```

What's the www_like count for males?
```
males <-pf[pf$gender == 'male' & !is.na(pf$gender),]
sum(males$www_likes)
females <- pf[pf$gender == 'female' & !is.na(pf$gender),]
sum(females$www_likes)

by(pf$www_likes, pf$gender, sum)
#pf$gender: female
#[1] 3507665
#------------------------------------------------------------------------------------------------------- 
#pf$gender: male
#[1] 1430175
```


***

### Box Plots
Notes:
http://flowingdata.com/2008/02/15/how-to-read-and-use-a-box-and-whisker-plot/
IQR (interquartile range): https://en.wikipedia.org/wiki/Interquartile_range

```{r Box Plots}
ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) +
  geom_boxplot()

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) +
  # ylim drops ovbservations
  #ylim(0, 1000) +    
  coord_cartesian(ylim = c(0, 1000)) +
  geom_boxplot()

by(pf$friend_count, pf$gender, summary)
```


***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
by(pf$friendships_initiated, pf$gender, summary)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)
mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
summary(pf$mobile_check_in)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

# % of users that checked_in with mobile
sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in) * 100
```

Response:
65%
***

### Analyzing One Variable
Reflection:

-  grab the initial idea of the data with summary, str, head and other functions to have a feeling of the data
-  understand types, distribution and anomalies of your variables
-  experiment with breakdowns - does the variable differ when break down by date or gender?
-  try to find out where the anomalies are coming from? is it just a single, extreme data point or is there a bigger story behind?
-  make sure numbers and plots you create do answer the question. Change the axis, scales, limits or plot type to get stuff done
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!